---
title: "Conditionals and Loops"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Conditionals

Conditionals let us make data-dependent choices about what to do next.

If
```{}
if (condition is true) {
  perform action
}
```

If else
```{}
if (condition is true) {
  do this 
} else {
  do that
}
```

Example
```{r}

x <- 5

if ( x > 10) {
  print("x is greater than 10") 
} else if (x > 5) {
  print("x > 5, but less than 10")
} else {
  print("x is 5 or less")

}
```

ifelse() for vector comparisons

```{r}
ifelse(condition, true action, false action)
```

```{r}

x <- 1:10
ifelse(x > 5, "x is greater than 5", "x is less than five")
```

# Looping in R

If you have a list/vector of things you want to do something with, looping is one way to do that. 

for loops say: for (iterator in list of elements), do this

```{r}
for (i in 1:10 ) {
  print(i)
}
```

Nested for loops - iterate over two things at once

```{r}
output <- character(10)
for (i in 1:5) {
  for (j in letters[1:5]) {
    output <- c(output, paste(i,j))
  }
}
```

```{r}
x
```

# Abstraction

What is R doing "behind the scenes" when we ask it to do this:

```{r}
system.time(replicate(100000,
1:50 + 1
))
```
It's looping! BUT, R knows how to do this in the most efficient way possible. Here's an equivalent for loop:

```{r}
add <- function(x) {
  y <- integer(x)
  for (i in 1:x) {
    y[i] <- i + 1
  }
  y
}


system.time(replicate(100000,
add(50)
))
```
```{r}
gapminder <- read.csv("../data/gapminder_data.csv")

(gapminder[(gapminder$year == 2002),]

rows2002 <- nrow(gapminder[(gapminder$year == 2002),])
rows2002

rows2002 >= 1

if (nrow(gapminder[(gapminder$year == 2002),]) >= 1) {
  print("records for 2002 found")
}

```

